Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-11-09 paper/paper.Rmd
move figure generation to separate files
Rmarkdown does less lifting

\begin{figure}
\caption{Replication of B\&S Figure 1 with both smooth and non-smoothed ROC curves\label{fig:rocs}}
\centering
\includegraphics[width=.95\linewidth]{figures/figure1-replicated.png}
\end{figure}

@@

```{r benefit-plot, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2). The Escalation model is the only model which consistently benefits from using smoothed ROC curves, and this advantage accounts for its apparent superiority to the other models."}
table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
table2_benefit <- read_csv("data/table2-smooth-benefit.csv")

cols <- c(
  Escalation = "#e41a1c",
  Goldstein = "#ff7f00",
  Quad = "#984ea3",
  CAMEO = "#377eb8",
  Average = "#4daf4a",
  `With PITF Predictors` = "gray50", 
  `Weighted by PITF` = "gray50", 
  `PITF Split Population` = "gray50",
  `PITF Only` = "gray50"
)

benefit <- bind_rows(
  table1_benefit %>%
    rename(setting = Model) %>%
    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff"),
  table2_benefit %>%
    rename(setting = Model, Escalation = `Escalation Only`) %>%
    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff")
)

# mean benefit by model
means <- benefit %>%
  group_by(model) %>%
  summarize(avg_benefit = mean(diff))

ann_text <- tibble(horizon = "1 month", diff = 0.07, model = 8.6, label = "Mean")

model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average", 
                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
                  "PITF Only")
benefit %>%
  mutate(model = factor(model, levels = rev(model_levels))) %>%
  ggplot(aes(y = model)) + 
  #facet_wrap(~ horizon, scales = "free_x") +
  geom_point(aes(x = diff, color = model), alpha = 0.5, size = 2) +
  scale_color_manual(guide = FALSE, values = cols) +
  geom_point(data = means, aes(x = avg_benefit), shape = "|", color = "black", size = 6) +
  labs(x = "AUC(smoothed ROC) - AUC(empirical ROC)",
       y = "") +
  theme_bw() +
  geom_text(data = ann_text, aes(x = diff, label = label), col = "black") +
  geom_vline(xintercept = 0, linetype = 3) +
  theme(axis.text.y = element_text(size = 12))
```

@@

The intention is to train seperate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest. Actually both random forests are trained on identical data, essentially production a random forest that just has N_trees \* 2.

@@

@@

```{r table1-plot, fig.height=5, fig.width=8, out.width = ".9\\linewidth", fig.align = "center", fig.cap = "Escalation to alternative comparisons for the ICEWS models (B\\&S Table 1)"}
table1_nosmooth <- read_csv("data/table1-nosmooth.csv")

icews_lvls <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average")

df <- table1_nosmooth %>%
  rename(Spec = Model) %>%
  pivot_longer(Escalation:Average, names_to = "Model", values_to = "auc_roc") %>%
  mutate(Model = factor(Model, levels = icews_lvls)) %>%
  nest(data = c(Spec, auc_roc))

pairs <- full_join(
  df %>% filter(Model!="Escalation"),
  df %>% filter(Model=="Escalation") %>% rename(data2 = data) %>% select(horizon, data2),
  by = c("horizon")
) %>%
  pivot_longer(data:data2, names_to = "xname", values_to = "data") %>%
  mutate(xname = ifelse(xname=="data", as.character(Model), "Escalation"),
         xname = factor(xname, levels = icews_lvls))

pairs %>%
  unnest(data) %>%
  ggplot(aes(x = xname, y = auc_roc)) +
  facet_grid(horizon ~ Model, scales = "free_x") +
  geom_point() +
  geom_line(aes(group = Spec)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(x = "", y = "AUC-ROC (empirical)")
```

```{r table2-plot, fig.height=5, fig.width=8, out.width = ".9\\linewidth", fig.align = "center", fig.cap = "Escalation to alternative comparisons for the structural models (B\\&S Table 2)"}
table2_nosmooth <- read_csv("data/table2-nosmooth.csv")

model_lvls <- colnames(table2_nosmooth)[-c(1:2)]

df <- table2_nosmooth %>%
  rename(Spec = Model) %>%
  pivot_longer(`Escalation Only`:`PITF Only`, names_to = "Model", values_to = "auc_roc") %>%
  mutate(Model = factor(Model, levels = model_lvls)) %>%
  nest(data = c(Spec, auc_roc))

pairs <- full_join(
  df %>% filter(Model!="Escalation Only"),
  df %>% filter(Model=="Escalation Only") %>% rename(data2 = data) %>% select(horizon, data2),
  by = c("horizon")
) %>%
  pivot_longer(data:data2, names_to = "xname", values_to = "data") %>%
  mutate(xname = ifelse(xname=="data", as.character(Model), "Escalation Only"),
         xname = factor(xname, levels = model_lvls))

pairs %>%
  unnest(data) %>%
  ggplot(aes(x = xname, y = auc_roc)) +
  facet_grid(horizon ~ Model, scales = "free_x") +
  geom_point() +
  geom_line(aes(group = Spec)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(x = "", y = "AUC-ROC (empirical)")
```

@@

```{r smooth-benefit-plot-extended, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2), with 3 different smoothing methods (binormal is the default method and used in B\\&S)"}

suppressPackageStartupMessages({
  library(tidyverse)
  library(yardstick)
  library(pROC)
  library(forcats)
})

preds <- read_rds("rep_nosmooth/output/all-predictions.rds")

auc_roc_vec <- function(pred, truth, smooth, sm = "binormal") {
  roc_obj <- pROC::roc(truth, pred, auc = TRUE, quiet = TRUE, smooth = smooth, smooth.method = sm)
  as.numeric(roc_obj$auc)
}

rocs <- preds %>% 
  filter(!is.na(value)) %>%
  group_by(cell_id) %>% 
  summarize(
    no = auc_roc_vec(pred = pred, truth = value, smooth = FALSE),
    binormal = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "binormal"),
    density = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "density"),
    fitdistr = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "fitdistr")
  )

# these two take a long time and error out for some reason
#logcondens = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "logcondens"),
#logcondens.smooth = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "logcondens.smooth")

mt <- read_rds("rep_nosmooth/output/model-table-w-results.rds")
mt <- mt %>% left_join(rocs, by = "cell_id")

cols <- c(
  Escalation = "#e41a1c",
  Goldstein = "#ff7f00",
  Quad = "#984ea3",
  CAMEO = "#377eb8",
  Average = "#4daf4a",
  `With PITF Predictors` = "gray50", 
  `Weighted by PITF` = "gray50", 
  `PITF Split Population` = "gray50",
  `PITF Only` = "gray50"
)

model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average", 
                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
                  "PITF Only")

benefit <- mt %>% 
  mutate(column = fct_recode(column, Escalation = "Escalation Only"),
         column = factor(column, levels = rev(model_levels))) %>%
  mutate(binormal = binormal - no,
         density  = density - no,
         fitdistr = fitdistr - no) %>%
  select(cell_id:non_RF, binormal:fitdistr) %>%
  pivot_longer(binormal:fitdistr, names_to = "method")

means <- benefit %>%
  group_by(column, method) %>%
  summarize(avg_benefit = mean(value))

ggplot(benefit, aes(x = value, y = column, color = column)) +
  facet_wrap(~method) +
  geom_point() +
  scale_color_manual(guide = FALSE, values = cols) +
  theme(axis.text.y = element_text(size = 12)) +
  geom_point(data = means, aes(x = avg_benefit), shape = "|", color = "black", size = 6) +
  labs(x = "AUC(smoothed ROC) - AUC(empirical ROC)",
       y = "") +
  theme_bw() +
  geom_vline(xintercept = 0, linetype = 3) +
  theme(axis.text.y = element_text(size = 12))
```

>>>>

\begin{figure}
\caption{Replication of B\&S Figure 1 with both smooth and non-smoothed ROC curves\label{fig:rocs}}
\centering
\includegraphics[width=.95\linewidth]{figures/fig-1-figure1-replicated.png}
\end{figure}

@@

\begin{figure}
\caption{Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\&S Tables 1 and 2). The Escalation model is the only model which consistently benefits from using smoothed ROC curves, and this advantage accounts for its apparent superiority to the other models.\label{fig:benefit-plot}}
\centering
\includegraphics[width=.8\linewidth]{figures/fig-2-benefit-plot.png}
\end{figure}

@@

The intention is to train separate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest. Actually both random forests are trained on identical data, essentially production a random forest that just has N_trees \* 2.

@@

\clearpage

@@

\begin{figure}
{\centering
\includegraphics[width=.9\linewidth]{figures/fig-A1-table1-pairs.png}
}
\caption{Escalation to alternative comparisons for the ICEWS models (B\&S Table 1)\label{fig:table1-plot}}
\end{figure}

\begin{figure}
{\centering
\includegraphics[width=.9\linewidth]{figures/fig-A2-table2-pairs.png}
}
\caption{Escalation to alternative comparisons for the structural models (B\&S Table 2)\label{fig:table2-plot}}
\end{figure}

@@

\begin{figure}
{\centering \includegraphics[width=.95\linewidth]{figures/fig-a3-benefit-plot-extended.png}}
\caption{Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\&S Tables 1 and 2), with 3 different smoothing methods (binormal is the default method and used in B\&S)\label{fig:smooth-benefit-plot-extended}}
\end{figure}
